---
title: ggplot2 compatible Quantile-quantile plots in R
author:
  - name: Alexandre Almeida
    affiliation: University of Campinas
    address:
    - Institute of Computing
    - Campinas, Brazil 13083-852
    email:  almeida.xan@gmail.com
  - name: Adam Loy
    affiliation: Carleton College
    address:
    - Department of Mathematics and Statistics
    - Northfield, MN 55057
    email:  aloy@carleton.edu
  - name: Heike Hofmann
    affiliation: Iowa State University
    address:
    - Department of Statistics
    - Ames, IA 50011-1210
    email:  hofmann@iastate.edu
abstract: >
  Univariate distributional assessment is a common thread throughout statistical analyses during both the exploratory and confirmatory stages. Q-Q plots compare two distributions by matching a commmon set of quantiles. Additional graphical elements are often added to Q-Q plots in order to aid in distributional assessment. Different orientations of Q-Q plots have also been proposed, most notably the de-trended Q-Q plot. Various implementations of Q-Q plots exist in R. \pkg{qqplotr} extends \pkg{ggplot2} to provide a complete implementation of Q-Q plots. In the remainder of this paper, we will introduce the plotting framework provided by \pkg{qqplotr} and provide multiple examples of how it can be used. 
preamble: >
  % Any extra latex you need in the preamble
  \usepackage{nameref}
  \usepackage{color}
output: rticles::rjournal_article
---

\newcommand{\hh}[1]{{\textcolor{orange}{#1}}}
\newcommand{\al}[1]{{\textcolor{violet}{#1}}}
\newcommand{\alex}[1]{{\textcolor{green}{#1}}}

## TODO:

* Abstract
* Conclusion

* Be consistent with indentation rules for code. We need to decide on these.
* Be consistent with colors/themes/sizes for the plots.
* Are we saying "normal" or "Normal" throughout the text?
* Do we want to use `bandType = "bs"` or should we change it to `bandType = "boot"`?
* Are we saying "color" or "colour" for ggplot2 code?

<!-- Loading supplementary packages to correctly draw the plots -->
```{r message=FALSE, warning=FALSE, echo=FALSE}
library(gridExtra)
library(tidyverse)
library(knitr)
knitr::opts_current$set(tidy=FALSE)
```

## Background
\label{sec:background}

Univariate distributional assessment is a common thread throughout statistical analyses during both the exploratory and confirmatory stages. When we begin exploring a new data set we often consider the distribution of individual variables before moving on to explore multivariate relationships. After a model has been fit to a data set, we must assess whether the distributional assumptions made are reasonable, and if they are not, then we must understand the impact this has on the conclusions of the model. Graphics provide arguably the most common way to carry out these univariate assessments. While there are many plots that can be used for distributional exploration and assessment, a quantile-quantile (Q-Q) plot \citep{Wilk1968-ii} is one of the most common plots used. 

Q-Q plots compare two distributions by matching a commmon set of quantiles. 
<!-- For two samples of size $n$, $X_1, X_2, \ldots, X_n$ and $Y_1, Y_2, \ldots, Y_n$, a Q-Q plot is a scatterplot comparing the sample quantiles.  -->
To compare a sample, $y_1, y_2, \ldots, y_n$ to a theoretical distribution, a Q-Q plot is simply a scatterplot of the sample quantiles, $y_{(i)}$, against the corresponding quantiles from the theoretical distribution, $F^{-1}\left( F_n(y_i) \right)$. If the empirical distribution is consistent with the theoretical distribution, then the Q-Q plot will be linear. For example, Figure \ref{fig:ex-qq1} shows two Q-Q plots: the left plot compares a sample drawn from a lognormal distribution to a lognormal distribution, while the right plot compares a sample drawn from a lognormal distribution to a normal distribution. As expected, the lognormal Q-Q plot is approximately linear, as the data and model are in agreement, while the normal Q-Q plot is curved, indicating disagreement between the data and the model.

```{r ex-qq1, echo=FALSE, out.width='.4\\linewidth', fig.width=3.5, fig.height=3.5, fig.show='hold', fig.align='center', fig.cap="The left plot compares a sample drawn from a lognormal distribution to a lognormal distribution, while the right plot compares a sample drawn from a lognormal distribution to a normal distribution. The curvature in the normal Q-Q plot highlights the disagreement betweeen the data and the model.", message = FALSE}
library(qqplotr)
set.seed(952017)
rand_sample <- data.frame(rs = rlnorm(20, meanlog = 0, sdlog = 1))

ggplot(data = rand_sample, mapping = aes(sample = rs)) +
  stat_qq_point(distribution = "lnorm") +
  ylab("Sample quantiles") +
  xlab("Theoretical quantiles") +
  ggtitle("Lognormal Q-Q plot") +
  theme_minimal()

ggplot(data = rand_sample, mapping = aes(sample = rs)) +
  stat_qq_point() +
  ylab("Sample quantiles") +
  xlab("Theoretical quantiles") +
  ggtitle("Normal Q-Q plot") +
  theme_minimal()
```

Additional graphical elements are often added to Q-Q plots in order to aid in distributional assessment. A reference line is often added to a Q-Q plot to help detection of departures from the proposed model.  This line is often drawn by connecting the first and third quartiles. Pointwise or simultaneous confidence bands are also frequently built around the reference line to display the expected degree of sampling error for the proposed model. Such bands hep gauge how troubling a departure from the proposed model may be.
Figure \ref{fig:ex-qq2} adds such reference lines and 95% pointwise confidence bands to the Q-Q plots in Figure \ref{fig:ex-qq1}.

```{r ex-qq2, echo=FALSE, out.width='.4\\linewidth', fig.width=3.5, fig.height=3.5, fig.show='hold', fig.align='center', fig.cap="Adding reference lines and $95\\%$ pointwise confidence bands to the Q-Q plots in Figure 1.", message = FALSE}
library(qqplotr)
set.seed(952017)
rand_sample <- data.frame(rs = rlnorm(20, meanlog = 0, sdlog = 1))

ggplot(data = rand_sample, mapping = aes(sample = rs)) +
  stat_qq_point(distribution = "lnorm") +
  stat_qq_line(distribution = "lnorm", color = "blue") +
  stat_qq_band(distribution = "lnorm", alpha = 0.5) +
  ylab("Sample quantiles") +
  xlab("Theoretical quantiles") +
  ggtitle("Lognormal Q-Q plot") +
  theme_minimal()

ggplot(data = rand_sample, mapping = aes(sample = rs)) +
  stat_qq_point() +
  stat_qq_line(color = "blue") +
  stat_qq_band(alpha = 0.5) +
  ylab("Sample quantiles") +
  xlab("Theoretical quantiles") +
  ggtitle("Normal Q-Q plot") +
  theme_minimal()
```

Different orientations of Q-Q plots have also been proposed, most notably the de-trended Q-Q plot. To detrend a Q-Q plot, the $y$-axis is changed to show the difference between the observed quantile and the reference line. Consequently, the line representing the agreement with the theoretical distribution is the $x$-axis. \citet{Loy2016-fg} find that detrended Q-Q plots are more powerful than other designs, \al{so long as $x$- and $y$-axes to show the same range. Adjusting the aspect ratio in this way ensures that distances in the $x$- and $y$-directions are on the same scale. This Q-Q plot design is called an *adjusted-detrended Q-Q plot*.  If the range of the axes are not adjusted, then this results an ordinary detrended Q-Q plot, which which was found to have lower power than the standard Q-Q plot in some situations \citep{Loy2016-fg}.} Figure \ref{fig:ex-detrend} displays the normal Q-Q plot from Figure \ref{fig:ex-qq2} along with its adjusted detrended version.

```{r ex-detrend, echo=FALSE, out.width='.4\\linewidth', fig.width=3.5, fig.height=3.5, fig.show='hold', fig.align='center', fig.cap="The left plot displays a traditional normal Q-Q plot for data simulated from a lognormal distribution. The right plot displays an adjusted detrended Q-Q plot of the same data, created by plotting the differences between the sample quantiles and the proposed model on the $y$-axis.", message = FALSE}
library(qqplotr)
set.seed(952017)
rand_sample <- data.frame(rs = rlnorm(20, meanlog = 0, sdlog = 1))

ggplot(data = rand_sample, mapping = aes(sample = rs)) +
  stat_qq_point() +
  stat_qq_line(color = "blue") +
  stat_qq_band(alpha = 0.5) +
  ylab("Sample quantiles") +
  xlab("Theoretical quantiles") +
  ggtitle("Normal Q-Q plot") +
  theme_minimal()

ggplot(data = rand_sample, mapping = aes(sample = rs)) +
  stat_qq_point(detrend = TRUE) +
  stat_qq_line(detrend = TRUE, color = "blue") +
  stat_qq_band(detrend = TRUE, alpha = 0.5) +
  ylab("Differences") +
  xlab("Theoretical quantiles") +
  ggtitle("Adjusted detrended version") +
  theme_minimal() +
  coord_fixed(ratio=1)
```

Various implementations of Q-Q plots exist in R. Normal Q-Q plots, where a sample is compared to the standard normal distribution, are implemented using `qqplot` and `qqline` in \pkg{base} graphics \citep{R}. \pkg{lattice} provides a general framework for Q-Q plots in the `qqmath` function, allowing one to compare a sample to any theoretical distribution by specifying the appropriate quantile function \citep{lattice}. `qqPlot` in the \pkg{car} package also allows for the assessment of non-normal distributions and adds pointwise confidence bands via normal theory or the parametric bootstrap \citep{car}. \pkg{ggplot2} provides `geom_qq` and `geom_qq_line`, enabling the creation of Q-Q plots with a reference line, much like those created using `qqmath` \citep{ggplot2}. None of these general use packages allow for easy construction of de-trended Q-Q plots.

\pkg{qqplotr} extends \pkg{ggplot2} to provide a complete implementation of Q-Q plots. The package allows for quick construction of all Q-Q plot designs without sacrificing the flexibility of the \pkg{ggplot2} framework. In the remainder of this paper, we will introduce the plotting framework provided by \pkg{qqplotr} and provide multiple examples of how it can be used. 

<!-- XXX `qualityTools` \cite{qualityTools} -->

## Implementing Q-Q plots in the \pkg{ggplot2} framework
\label{sec:implementing}

<!-- With \pkg{qqplotr} we extend some of the original \pkg{ggplot2} Q-Q plot  -->
<!-- functionatilites by permitting the drawing of Q-Q points, lines, and confidence  -->
<!-- bands.  -->
\pkg{qqplotr} provides a \pkg{ggplot2} layering mechanism for Q-Q points, lines, and confidence bands by implementing separate statistical transformations (`stat`s). In this section, we describe each transformation.

<!-- The Q-Q plot functions are divided into three statistical transformations.  -->
<!-- Below, we describe each one of those functions and also give an overview of its -->
<!-- parameters. -->

### `stat_qq_point`

This is a modified version of `stat_qq`/`geom_qq` from \pkg{ggplot2} that plots 
the sample quantiles versus the theoretical quantiles (as in Figure 
\ref{fig:ex-qq1}). The novelty of this implementation is an option to detrend 
the plotted points (see \nameref{sec:background}). Note that all other 
implemented functions in the \pkg{qqplotr} package also allow for this option.
Below we present a complete call to `stat_qq_point` and highlight its default
parameters values:

```{r echo=TRUE, eval=FALSE}
stat_qq_point(
  data = NULL,
  mapping = NULL,
  geom = "point",
  position = "identity",
  na.rm = TRUE,
  show.legend = NA,
  inherit.aes = TRUE,
  distribution = "norm",
  dparams = list(),
  detrend = FALSE,
  qtype = 7,
  qprobs = c(0.25, 0.75),
  ...
  )
```

* Parameters such as `data`, `mapping`, `geom`, `position`, `na.rm`, 
`show.legend`, and `inherit.aes` are specific for \pkg{ggplot2} implementations.

* `distribution` is a character string that sets the theoretical probability 
distribution. Here, we followed the nomenclature from the \pkg{stats} package,
but rather than requiring the full function name for a distribution (e.g., `"dnorm"`), only the suffixe is required (e.g., `"norm"`).
If you wish to provide a
custom distribution, then you must first create density (PDF), distribution (CDF), quantile, and simulation
functions, following nomenclature outlined in the \pkg{stats} package (i.e., for
`"custom"`, you must provide the `dcustom`, `pcustom`, `qcustom`, and 
`rcustom` functions). A detailed example is given in the \nameref{sec:user-dists} section.

* `dparams` is a named list specifying the parameters of the proposed `distribution`.
By default, maximum likelihood etimates (MLEs) are used, so this argument overrides this default.
Please note that MLEs are currently only supported for
 distributions available in the \pkg{stats} package, so if a  custom distribution is used in `distribution` *all* of its parameters have to be provided in `dparams`.

* `detrend` is a boolean that controls whether the points should be detrended (as 
shown in Figure \ref{fig:ex-detrend}). More on that in \nameref{sec:detrending}.

* `qtype` and `qprobs` are only used when `detrend = TRUE`. These parameters are
passed on to the `type` and `probs` parameters of the `quantile` function from the
\pkg{stats} package, which is used to define the quantiles the reference line (i.e., the proposed model) passes through.


### `stat_qq_line`

This statistical transformation draws a reference line based on the sample data 
quantiles. Note that the parameters for `stat_qq_line` are identical to those for `stat_qq_point`.

```{r echo=TRUE, eval=FALSE}
stat_qq_line(
  data = NULL,
  mapping = NULL,
  geom = "path",
  position = "identity",
  na.rm = TRUE,
  show.legend = NA,
  inherit.aes = TRUE,
  distribution = "norm",
  dparams = list(),
  detrend = FALSE,
  qtype = 7,
  qprobs = c(0.25, 0.75),
  ...
  )
```

Two lines are of particular importance in a Q-Q plot:

a) By definition of a Q-Q plot the *identity line* represents the theoretical distribution. 
b) The *Q-Q line* is an estimate of the empirical distribution. 

The intercept and slope of these lines relate to mean and scale parameters of the theoretical and the empirical distribution, respectively.
By comparing these two lines we learn about how well the  parameters estimated from the sample match the theoretical  parameters.  For a distributional family that is invariant to linear transformations, the parameters specified in the theoretical distribution only have an effect on the Q-Q line and the Q-Q points---that is, the parameters get shifted and scaled in the plot, but relative relationships do not change aside from a change in scales.  For other distributions, such as a lognormal distribution, re-specifications of the parameters result in non-linear transformations of the Q-Q line and Q-Q points (see Figure \ref{fig:qqline} for an example).  


```{r qqline, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=8, out.width='\\textwidth', fig.cap="Q-Q plots of (top-left) a mis-specified Normal distribution, (top-right)  a Normal distribution with ML parameter estimates. Between these two plots only the scales on the axes change. The bottom two plots show Q-Q plots of the same sample from a log-normal distribution. On the left, the theoretical distribution is mis-specified with a mean of 0 and variance of 1. On the right, ML estimates for mean and variance are used."}
library(tidyverse)
library(qqplotr)
library(gridExtra)

dframe <- data.frame(xnorm = rnorm(50, mean=10, sd=2),
           xlnorm = rlnorm(50, mean=2, sd=.5))
p11 <- dframe %>% ggplot(aes(sample=xnorm)) + 
  geom_qq_band(dparams=list(mean=0, sd=1)) + 
  stat_qq_line(dparams=list(mean=0, sd=1)) + 
  stat_qq_point(dparams=list(mean=0, sd=1)) +
  theme_bw() +
  xlab("Theoretical Quantiles") +
  ylab("Observed Quantiles") +
  ggtitle("Normal Distribution N(0,1)")
p12 <- dframe %>% ggplot(aes(sample=xnorm)) + 
  geom_abline() +
  geom_qq_band() + 
  stat_qq_line() + 
  stat_qq_point() +
  theme_bw() +
  xlab("Theoretical Quantiles") +
  ylab("Observed Quantiles") +
  ggtitle(expression("Normal Distribution N("~mu~","~sigma~")"))
p21 <- dframe %>% ggplot(aes(sample=xlnorm)) + 
  geom_qq_band(distribution="lnorm", dparams=list(mean=0, sd=1)) + 
  stat_qq_line(distribution="lnorm", dparams=list(mean=0, sd=1)) + 
  stat_qq_point(distribution="lnorm", dparams=list(mean=0, sd=1)) +
  theme_bw() +
  xlab("Theoretical Quantiles") +
  ylab("Observed Quantiles") +
  ggtitle(expression("log Normal Distribution LN(0,1)"))
p22 <- dframe %>% ggplot(aes(sample=xlnorm)) + 
  geom_abline() +
  geom_qq_band(distribution="lnorm") + 
  stat_qq_line(distribution="lnorm") + 
  stat_qq_point(distribution="lnorm") +
  theme_bw() +
  xlab("Theoretical Quantiles") +
  ylab("Observed Quantiles") +
  ggtitle(expression("log Normal Distribution LN("~mu~","~sigma~")"))

grid.arrange(p11,p12,p21,p22, nrow=2, ncol=2)
```

\hh{By default the Q-Q line is drawn through two points, corresponding to the .25 and .75 quantiles of the distributions on $x$ (theoretical distribution) and $y$ (empirical distribution).  While we are using MLEs to identify the theoretical distribution, these estimates are only used in determining the $x$-coordinates of these points. For the $y$ coordinates, the 1st and 3rd quartiles are used by default for more robust estimation. Therefore, the difference between the observed and theoretical distributions presents itself in the difference between the Q-Q line and the identity line. Using robust estimates for the Q-Q line is of particular advantage for small samples \citep{Loy2016-fg}.}




### `stat_qq_band`

Confidence bands can be drawn around the reference line using one of three methods: a 
normal approximation, the parametric bootstrap, or the tail-sensitive procedure \citep{Aldor-Noiman2013-xw}.
Below, we present the call to `stat_qq_band` and describe its specific 
parameters:

```{r echo=TRUE, eval=FALSE}
stat_qq_band(
  data = NULL,
  mapping = NULL,
  geom = "qq_band",
  position = "identity",
  show.legend = NA,
  inherit.aes = TRUE,
  na.rm = TRUE,
  distribution = "norm",
  dparams = list(),
  detrend = FALSE,
  qtype = 7,
  qprobs = c(0.25, 0.75),
  bandType = "normal",
  B = 1000,
  conf = 0.95,
  mu = NULL,
  sigma = NULL,
  ...
  )
```

* `bandType` is a character string controlling the method used to
construct the confidence bands:
    + **Normal:** Specifying `bandType = "normal"` (default) constructs pointwise confidence
bands based on the normal approximation to the distribution of the order
statistics. For example, an approximate 95\% confidence interval for the $i$th
order statistic is $\widehat{X}_{(i)}~\pm~\Phi^{-1}(.975)~\cdot~SE(X_{(i)})$,
where $\widehat{X}_{(i)}$ denotes the value along the fitted line,
$\Phi^{-1}(\cdot)$ denotes the quantile function for the standard normal
distribution, and $SE(X_{(i)})$ is the standard error of the $i$th order
statistic.
    + **Bootstrap:** Specifying `bandType = "bs"` constructs pointwise confidence 
bands using percentile confidence intervals from the parametric bootstrap.
    + **Tail-sensitive:** Specifying `bandType = "ts"` constructs the simulation-based
tail-sensitive simultaneous confidence bands proposed by
\citet{Aldor-Noiman2013-xw}.

* `B` is an integer spoecifying the number of bootstrap replicates (if
`bandType = "bs"`) or the number of simulated samples (if `bandType = "ts"`).

* `conf` is a numerical variable setting the confidence level of the bands,
which is bound between 0 and 1.

* `mu` and `sigma` are only used when `bandType = "ts"`. They represent the 
center and scale parameters, respectively, which are used to
construct the simulated tail-sensitive confidence bands. If any
parameters is `NULL`, then both the center and scale parameters
will be estimated using robust estimates via the `robustbase` package \citep{robustbase}.

### Groups in `qqplotr`

`qqplotr` is implemented in accordance with the `ggplot2` concept of groups. When the user maps values to aesthetics that explicitly (by using `group`) or implicitly (such as `shape` or discrete values of `color, size`, etc.) introduce groups, the corresponding calculations respect the grouping in the data. All the groups are compared to the same distributional family specified, but the parameters are *separately* estimated for each of the groups (that is, given that `dparams = NULL`).
If the user wants to fit the **same** distribution (i.e., same parameters estimates) to each group, then the estimates must be manually calculated and passed to `dparams` as a named list for each of the desired \pkg{qqplotr} functions.

<!-- In the case that the user wants to fix a particular distribution or set of distributions. This can be achieved by manually setting/calculating  estimators beforehand, storing them into a list, and providing this list to `dparams` of the appropriate \pkg{qqplotr} functions. -->

The use of groups is illustrated in more detail in the \nameref{sec:brfss} section.

\FloatBarrier

## Examples
\label{sec:examples}

In this section, we demonstrate the capabilities of the \pkg{qqplotr} package. We start by loading
the package:

```{r message=FALSE}
library(qqplotr)
```


### Constructing Q-Q plots with `qqplotr`

To give a brief introduction on how to use \pkg{qqplotr} and its functions, let us use as example the `urine` dataset from the \pkg{boot} package. This small dataset presents 79 urine specimens that were analyzed in order to determine if certain physical characteristics of the urine (e.g. pH or urea concentration) might be related to the formation of calcium oxalate crystals. In this example, we will focus on the distributional assessment of pH measurements made on the samples.

We start by plotting a normal Q-Q plot of the data. Figure \ref{fig:urine-qq} shows the Q-Q plot of the pH measurements versus the normal distribution quantiles. As previously described, the theoretical distributional paremeters are estimated on the fly with MLEs. By default, `stat_qq_band` creates pointwise confidence bands based on the normal approximation to the distribution of the order statistics (i.e., `bandType = "normal"`). As we can see, the urine pH measurements distribution for this dataset is lightly right-skewed. The chosen normal confidence bands suggests that the assumption of normality is not appropriate to the data.

```{r urine-qq, echo = TRUE, message = FALSE, warning = FALSE, fig.width = 3.5, fig.height = 3.5, fig.cap = "Normal Q-Q plot comparing pH measurements of the urine samples with a normal distribution. \\al{Do we really need this plot? We could reference Figure 6 as easily.}", out.width='0.45\\textwidth', fig.align="center"}
require(boot)
urine %>% ggplot(aes(sample = ph)) + 
  stat_qq_band(alpha = 0.5) + 
  stat_qq_line() + 
  stat_qq_point() +
  xlab("Theoretical quantiles") +
  ylab("Observed quantiles of pH") +
  theme_bw()
```

\FloatBarrier

As previously discussed, three distinct approaches are availble in \pkg{qqplotr} to construct the confidence bands. 
<!-- For instance, \citet{Aldor-Noiman2013-xw} discuss that their proposed tail-sensitive simultaneous confidence bands offers more statistical power, especially when distinguishing normal distributions from mildy or heavy-tailed distributions. -->
Figure \ref{fig:urine-qq-bands} presents three plots with the different confidence bands overlayed on the same normal Q-Q points and reference lines. Notice that the normal-based confidence bands (on the left) are the same as in Figure \ref{fig:urine-qq}. When assessing the bootstrap (center) and tail-sensitive confidence bands (on the right), we can see that more points are inside the confidence bands than with the normal confidence bands method. Hence, one could conclude by looking at the tail-sensitive plot at the right hand side that, even though the normal distribution seems like a rough approximation, the assumption of normality is reasonable for the urine pH measurements.

```{r urine-qq-bands, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 9.5, fig.height = 3.5, fig.cap = "Normal Q-Q plots of pH measurements from urine samples using different confidence bands. Depending on the type of confidence band used, we come to different conclusions.", out.width='1.0\\textwidth', fig.align="center"}
p9 <- urine %>% ggplot(aes(sample = ph)) + 
  stat_qq_band(alpha = 0.5, bandType = "normal", fill = "#66C2A5") + 
  stat_qq_line(col = "#66C2A5") + #"gray40") + 
  stat_qq_point() +
  ggtitle("Normal") +
  xlab("Theoretical quantiles") +
  ylab("Observed quantiles of pH") +
  theme_bw() +
  coord_cartesian(ylim = c(3.2, 8.7))

p10 <- urine %>% ggplot(aes(sample = ph)) + 
  stat_qq_band(alpha = 0.5, bandType = "bs", fill = "#FC8D62") + 
  stat_qq_line(col = "#FC8D62")+ #"orange2") + 
  stat_qq_point() +
  ggtitle("Bootstrap") +
  xlab("Theoretical quantiles") +
  ylab("Observed quantiles of pH") +
  theme_bw() +
  theme(axis.text.y = element_blank(), 
          axis.ticks.y = element_blank(), 
          axis.title.y = element_blank(),
          plot.title = element_text(vjust = 0),
          plot.margin = margin(t = 3.5, r = 5.5, b = 5.5, l = 5.5, unit = "pt")) +
  coord_cartesian(ylim = c(3.2, 8.7))

p11 <- urine %>% ggplot(aes(sample = ph)) + 
  stat_qq_band(alpha = 0.5, bandType = "ts", fill = "#8DA0CB") + 
  stat_qq_line(col = "#8DA0CB") + #"steelblue") + 
  stat_qq_point() +
  ggtitle("Tail-sensitive") +
  xlab("Theoretical quantiles") +
  ylab("Observed quantiles of pH") +
  theme_bw() +
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank(), 
        axis.title.y = element_blank()) +
  coord_cartesian(ylim = c(3.2, 8.7))

grid.arrange(p9, p10, p11, ncol = 3, widths = c(1.11, 1, 1))
```

\FloatBarrier

### User-provided distributions
\label{sec:user-dists}

Using the capabilities of \pkg{qqplotr} with the distributions implemented in the \pkg{stats} package is relatively straightfoward, since the implementation allows you to specify the suffix (i.e., distribution and or abbreviation) via the `distribution` argument and the parameter estimates via the `dparams` argument. However, there are times when the distributions in \pkg{stats} are not sufficient for the demands of the analysis. For example, there is no left-skewed distribution listed. User-coded distributions, or distributions from other packages, can be used with \pkg{qqplotr} as long as the distributions are defined following the conventions laid out in the \pkg{stats} package. Specfically, for some distribution there must be density/mass (`d` prefix), CDF (`p` prefix), quantile (`q` prefix), and simulation (`r` prefix) functions. In this section, we illustrate the use of the smallest extreme value distribution (SEV).

To qualify for the 2012 Olympics in the men's long jump athletes had to meet/exceed the 8.1 meter standard or place in the top twelve. During the qualification events, each athlete was able to jump up to three times, using their best (i.e., longest) jump as the result. Figure \ref{fig:jump-density} shows a density plot of the results, which are clearly left skewed. 


```{r jump-density, echo=FALSE, fig.width = 3.5, fig.height = 3.5, fig.cap = "Density plot of the 2012 men's long jump qualifying round. The distances are clearly left skewed.", out.width='0.45\\textwidth', fig.align="center", message=FALSE, warning = FALSE}
longjump <- na.omit(longjump)
ggplot(longjump, aes(distance)) +
  geom_density(fill = "orange", color = "orange", alpha = 0.5) +
  xlab("Jump distance (in m)") +
  theme_bw()
```


In order to model the jump distances we must first define a left-skewed distribution. Below, we define the suite of distributional functions necessary to utilize the SEV distribution.

```{r}
# CDF
psev <- function(q, mu = 0, sigma = 1) {
    z <- (q - mu) / sigma
    1 - exp(-exp(z))
}

# PDF
dsev <- function(x, mu = 0, sigma = 1) {
  z <- (x - mu) / sigma
  (1 / sigma) * exp(z - exp(z))
}

# Quantile function
qsev <- function(p, mu = 0, sigma = 1) {
  mu + log(-log(1 - p)) * sigma
}

# Simulation function
rsev <- function(n, mu = 0, sigma = 1) {
  qsev(runif(n), mu, sigma)
}
```

With the `*sev` distribution functions in hand, we can create a Q-Q plot to assess the appropriateness of the SEV model (Figure \ref{fig:sev-qq}). The Q-Q plot shows that the distances do not substantially deviate from the SEV model, so we have found an adequate representation of the distances. The code used to create Figure \ref{fig:sev-qq} is given below:

```{r sev-qq, fig.cap = "Q-Q plot comparing the long jump distances to the standard SEV distribution with 95\\% pointwise confidence bands. The SEV distribution appears to adequately model the distances.", fig.width = 3.5, fig.height = 3.5, out.width='0.45\\textwidth', fig.align="center", message=FALSE, warning = FALSE}
ggplot(longjump, aes(sample = distance)) +
  stat_qq_band(distribution = "sev", dparams = list(mu = 0, sigma = 1), 
               alpha = 0.3) +
  stat_qq_line(distribution = "sev", dparams = list(mu = 0, sigma = 1)) +
  stat_qq_point(distribution = "sev", dparams = list(mu = 0, sigma = 1)) +
  xlab("Theoretical quantiles") +
  ylab("Jump distance (in m)") +
  theme_bw()
```

### Detrending Q-Q plots
\label{sec:detrending}


<!-- \al{AL: It seems natural to have a subsection here, even if we continue the example, to allow people to skim the paper and find the topic they are most interested in. I tried to eliminate the techinical discussion of the plot since it is in the background section, instead describing the code. What do you think?} -->

<!-- \pkg{qqplotr} also allows for Q-Q plots to be *detrended*. In a detrended Q-Q plot, the $y$-axis shows the difference between the empirical quantile and the reference line (i.e. the theoretical distribution). This layout directly plots what we want viewers to assess --the difference between the distributions being compared-- which \citet{Loy2016-fg} found to be more powerful than other designs, so long as the $y$-axis limits are set so that the aspect ratio is kept the same as in the traditional Q-Q plot. -->

To illustrate how to construct an adjusted detrended Q-Q plot using \pkg{qqplotr}, consider detrending Figure \ref{fig:sev-qq}. This is done by adding the argument `detrend = TRUE` to `stat_qq_point`, `stat_qq_line`, and `stat_qq_band`. To adjust the aspect ratio to ensure that vertical and horizontal distances are on the same scale we further add `coord_fixed(ratio=1)`. We leave it to the user to adjust the $y$-axis limits on a case-by-case basis. The full command to construct Figure \ref{fig:detrend-sev} is given below:


<!-- For example, Figure \ref{fig:detrend-sev} compares the standard Q-Q plot shown in Figure \ref{fig:sev-qq} with a detrended version by adding the argument `detrend = TRUE` to the `stat_qq_band`, `stat_qq_line`, and `stat_qq_point` calls. -->

```{r detrend-sev, echo=FALSE, fig.width = 3.5, fig.height = 3.5, out.width='0.45\\textwidth', fig.cap = "An adjusted detrended Q-Q plot assessing the appropriateness of the SEV distribution for the long jump data.", fig.align="center", message=FALSE, warning = FALSE}
p7 <- ggplot(longjump, aes(sample = distance)) +
  stat_qq_band(distribution = "sev", dparams = list(mu = 0, sigma = 1),
               alpha = 0.3) +
  stat_qq_line(distribution = "sev", dparams = list(mu = 0, sigma = 1)) +
  stat_qq_point(distribution = "sev", dparams = list(mu = 0, sigma = 1)) +
  xlab("Theoretical quantiles") +
  ylab("Jump distance (in m)") +
  theme_bw()

p8 <- ggplot(longjump, aes(sample = distance)) +

  stat_qq_band(distribution = "sev",  detrend = TRUE, dparams = list(mu = 0, sigma = 1), alpha = 0.3) +
  stat_qq_line(distribution = "sev", detrend = TRUE, dparams = list(mu = 0, sigma = 1)) +
  stat_qq_point(distribution = "sev", detrend = TRUE, dparams = list(mu = 0, sigma = 1)) +
  xlab("Theoretical quantiles") +
  ylab("Differences") +
  theme_bw() + 
  coord_fixed(ratio=1, ylim = c(-2, 2))

# grid.arrange(p7, p8, ncol = 2)
p8
```

```{r eval=FALSE}
ggplot(longjump, aes(sample = distance)) +
  stat_qq_band(distribution = "sev", detrend = TRUE, dparams = c(mu = 0, sigma = 1),  
               alpha = 0.3) +
  stat_qq_line(distribution = "sev", detrend = TRUE, dparams = c(mu = 0, sigma = 1)) +
  stat_qq_point(distribution = "sev", detrend = TRUE, dparams = c(mu = 0, sigma = 1)) +
  xlab("Theoretical quantiles") +
  ylab("Differences") +
  theme_bw() +
  coord_fixed(ratio=1)
```

<!-- In order to create a so-called *adjusted* detrended Q-Q plot \citep{Loy2016-fg} the aspect ratio must also be set to 1. If the aspect ratio is not adjusted in this way, an *ordinary* detrended Q-Q plot is created, which was found to have lower power than the standard Q-Q plot in some situations \citep{Loy2016-fg}. -->



### BRFSS example
\label{sec:brfss}

The Center for Disease Control and Prevention runs an annual telephone survey, the Behavioral Risk Factor Surveillance System (BRFSS), to keep track of the US populations' "health-related risk behaviors, chronic health conditions, and use of preventive services." \al{(We need a citation for this quote.)}  
Close to half a million interviews are conducted each year. In this example, we focus on the responses for Iowa in 2012. The data set consists of 7166 responses across 359 questions and derived variables. To further illustrate the functionality of \pkg{qqplotr}, we focus on assessing the distributions of Iowan's heights and weights.
<!-- Among these, are participants' heights and weights, which we are going to assess in more detail.  -->

Figure \ref{fig:heights} shows two Q-Q plots side by side. For each of the plots, a sample of 200 men and 200 women is drawn from the overall number of responses. On the left hand side, individuals' heights are plotted in a Q-Q plot comparing raw heights to a normal distribution. We see that the distributions for both men and women (colour) show horizontal steps: this indicates that the distributional assessement is heavily dominated by the discreteness in the data, as most respondents provided their height to the closest inch. On the right hand side of Figure \ref{fig:heights}, we use jittering to remedy this situation---that is, we add a random number generated from a random uniform distribution on $\pm 0.5$ inch to the reported height, as shown in the code below: \al{We need to add the code! Should we show the entire chunk below?}

```{r iowa-read, echo=12:14, message=FALSE, warning=FALSE}
iowa <- readRDS("data/brfss-iowa.rda")
set.seed(3145)

sample_ia <- iowa %>% nest(-SEX) %>% mutate(
  data = data %>% purrr::map(.f = function(x) sample_n(x, size = 200))  
) %>% unnest(data) %>% dplyr::select(SEX, WTKG3, HTIN4) %>%
  mutate(Gender = c("Male", "Female")[SEX])

params <- iowa %>% filter(!is.na(HTIN4)) %>% summarize(
  m = mean(HTIN4),
  s = sd(HTIN4)
)
customization <- list(scale_fill_brewer(palette="Set1"),
  scale_colour_brewer(palette = "Set1"),
  xlab("Theoretical quantiles"),
  ylab("Height (in inch)"),
  coord_equal(),
  theme_bw(),
  theme(legend.position = c(0.8, 0.2), aspect.ratio = 1))

p1 <- sample_ia %>% ggplot(aes(sample = HTIN4, colour=Gender, fill=Gender)) + 
  stat_qq_band(alpha = 0.3, dparams=list(mean=params$m, sd=params$s)) + 
  stat_qq_line( dparams=list(mean=params$m, sd=params$s)) + 
  stat_qq_point( dparams=list(mean=params$m, sd=params$s)) +
  customization

p2 <- sample_ia %>% mutate(
    HTIN4.jitter = jitter(HTIN4, factor = 2)
    ) %>% 
  ggplot(aes(sample = HTIN4.jitter, colour=Gender, fill=Gender)) + 
  geom_abline(slope=1, intercept=0, colour = "grey40") +
  stat_qq_band(alpha = 0.3, dparams = list(mean=params$m, sd=params$s)) +
  stat_qq_line(dparams = list(mean=params$m, sd=params$s)) + 
  stat_qq_point(dparams = list(mean=params$m, sd=params$s)) +
  customization +
  ylab("Jittered Height (in inch)") 
```

By using jittering we diminish the effect that discreteness has on the distribution and brings the observed distribution much closer to a normal distribution. Note that separate normal distributions were fit to each gender. Unsurprisingly, the resulting distributions have different means (women are, on average, 6 inches shorter than men in this dataset). Interestingly, the slope of the two genders is similar, indicating that the same scale parameter fits both genders' distributions (the standard deviation of height in the data set is 2.97 inch for men and 2.91 inch for women, see Table \ref{tab:heights}). 
The dark line between the two groups is the identity line, representing the theoretical distribution each group is compared to. This distribution is based on parameters estimated from the whole population (see Table \ref{tab:heights} for the estimates). While the mean is about half way between the gender means,  the standard deviation of the height based on the whole population is larger, as seen by the steeper slope of the identity compared to the lines for each group.

```{r heights-table, echo=FALSE, results="asis"}
bygender <- iowa %>% group_by(SEX) %>%
  summarize(
    mean.height = mean(HTIN4, na.rm=TRUE),
    sd.height = sd(HTIN4, na.rm=TRUE),
    mean.weight = mean(log(WTKG3), na.rm=TRUE),
    sd.weight = sd(log(WTKG3), na.rm=TRUE)
  )
bygender$SEX <- c("Male", "Female")

total <- iowa %>% 
  summarize(
    mean.height = mean(HTIN4, na.rm=TRUE),
    sd.height = sd(HTIN4, na.rm=TRUE),
    mean.weight = mean(log(WTKG3), na.rm=TRUE),
    sd.weight = sd(log(WTKG3), na.rm=TRUE)
  )

bygender[3,] <- c("Total", total)
names(bygender) <- c("SEX", "mean height (in inch)", "sd (in inch)", "mean log weight (in kg)", "sd (in kg)")
knitr::kable(bygender, format = "latex",  
             caption = "Summary of Iowa residents' heights and weights with corresponding standard deviations by gender and for the total population.\\label{tab:heights}", 
             digits=2, booktabs = TRUE)
```

```{r heights, echo=FALSE, fig.width = 8, fig.height = 3.5, fig.cap = "Q-Q plots comparing the raw (left) and jittered (right) heights to the normal distribution for a sample 200 men and 200 women. The distribution on the left is dominated by the discreteness of the data. On the right, using the normal distribution to model people's height is not completely absurd, except for a few extreme outliers.", out.width='\\textwidth', fig.align="center", message=FALSE, warning = FALSE}
grid.arrange(p1, p2, ncol = 2)
```

Unlike respondents' heights, their weights do not seem to be normally distributed. Figure \ref{fig:weights} shows two Q-Q plots of these data. The Q-Q plot on the left compares raw weights to a normal distribution. We see that tails of the observed distribution are heavier than expected under a normal distribution. On the right, weights are log-transformed. We see that a normal distribution for each of the genders appears to be reasonable, with the exception of a few extreme outliers.

```{r weights, echo=FALSE, fig.width = 8, fig.height = 3.5, fig.cap = "Q-Q plots comparing weights to the normal distribution for a sample of 200 men and 200 women. Unlike people's height, weight seems to be heavily right skewed with some additional outliers in the left tail (left plot). On the right, weight was log-transfomed before its distribution is compared to a theoretical normal. ", out.width='\\textwidth', fig.align="center", message=FALSE, warning = FALSE}
params <- iowa %>% filter(!is.na(WTKG3)) %>% summarize(
  m = mean(WTKG3/100),
  s = sd(WTKG3/100)
)
p3 <- sample_ia %>% 
  ggplot(aes(sample = WTKG3/100, colour=Gender, fill=Gender)) + 
  stat_qq_band(alpha = 0.3, dparams=list(mean=params$m, sd=params$s)) + 
  stat_qq_line(dparams=list(mean=params$m, sd=params$s)) + 
  stat_qq_point(dparams=list(mean=params$m, sd=params$s)) +
  scale_fill_brewer(palette="Set1") +
  scale_colour_brewer(palette = "Set1") +
  xlab("Theoretical quantiles") +
  ylab("Weight (in kg)") +
  coord_equal() +
  theme_bw() +
  theme(legend.position = c(0.8, 0.2), aspect.ratio = 1)

p5 <- sample_ia %>% 
  ggplot(aes(sample = log(WTKG3/100), colour=Gender, fill=Gender)) + 
  stat_qq_band(alpha = 0.3, dparams=list(mean=params$m, sd=params$s)) + 
  stat_qq_line( dparams=list(mean=params$m, sd=params$s)) + 
  stat_qq_point( dparams=list(mean=params$m, sd=params$s)) +
  scale_fill_brewer(palette="Set1") +
  scale_colour_brewer(palette = "Set1") +
  xlab("Theoretical quantiles") +
  ylab("log Weight (in kg)") +
  coord_equal() +
  theme_bw() +
  theme(legend.position = c(0.8, 0.2), aspect.ratio = 1)

# don't need the jittering
grid.arrange(p3, p5, ncol = 2)
```

Instead of log-transforming the observed weights, we can change the theoretical distribution to lognormal. 
Figure \ref{fig:weights-log} shows two lognormal Q-Q plots, one for each gender. note that by default, the MLEs are used to parameterize the lognormal distribution for each group:  

```{r weights-log, echo=6:9, fig.width = 8, fig.height = 3.5, fig.cap = "Q-Q plots comparing weights to a lognormal distribution for a sample of 200 men and 200 women. Shift and scale parameters are estimated separately for each of the genders before comparing distributions to a log-normal.", out.width='.9\\textwidth', fig.align="center", message=FALSE, warning = FALSE}

customization <- list(
  scale_fill_brewer(palette = "Set1"), 
  scale_colour_brewer(palette = "Set1"),
  xlab("Theoretical quantiles"),
  ylab("Weight (in kg)"),
  theme_bw(),
  theme(legend.position = "none", aspect.ratio = 1)
)

mlog <- mean(log(sample_ia$WTKG3 / 100), na.rm = TRUE)
sdlog <- sd(log(sample_ia$WTKG3 / 100), na.rm = TRUE)

p4 <- sample_ia %>% 
  ggplot(aes(sample = WTKG3 / 100, colour = Gender, fill = Gender)) +
  geom_abline(colour = "grey60") +
  stat_qq_band(
    distribution = "lnorm",
    dparams = list(meanlog = mlog, sdlog = sdlog),
    alpha = 0.3
  ) +
  stat_qq_line(
    distribution = "lnorm",
    dparams = list(meanlog = mlog, sdlog = sdlog)
  ) +
  stat_qq_point(
    distribution = "lnorm",
    dparams = list(meanlog = mlog, sdlog = sdlog)
  ) +
  customization +
  theme(legend.position = c(0.2, 0.8)) 

p6 <- sample_ia %>% 
  ggplot(aes(sample = WTKG3 / 100, colour = Gender, fill = Gender)) +
  geom_abline(colour="grey60") +
  stat_qq_band(distribution = "lnorm", alpha = 0.3) +
  stat_qq_line(distribution = "lnorm") +
  stat_qq_point(distribution = "lnorm") +
  customization + facet_grid(.~Gender)

#grid.arrange(p4, p6, ncol = 2)
p6
```


<!-- ### Q-Q plots with robust estimators -->

<!-- \al{What do we really want to highlight with robust estimators? I am ready to write this section, but am unsure what to highlight: the different lines? the dparams?} -->


```{r include=FALSE}
set.seed(798534)
t_data <- rt(20, df = 2)
norm_data <- rnorm(20)
chisq1 <- rchisq(20, df = 1)
```



```{r include=FALSE}
ggplot(data.frame(x = t_data), aes(sample = x)) +
  # stat_qq_band(distribution = "norm", alpha = 0.3) +
  stat_qq_line(distribution = "norm") +
  stat_qq_point(distribution = "norm") +
  geom_abline(intercept = mean(t_data), slope = sd(t_data), color = "orange") +
  geom_abline(intercept = median(t_data), slope = mad(t_data), color = "blue") +
  geom_abline(intercept = median(t_data), slope = robustbase::Qn(t_data), color = "green") +
  geom_abline(intercept = median(t_data), slope = robustbase::Sn(t_data), color = "red") +
  xlab("Theoretical quantiles") +
  xlab("Sample quantiles")

ggplot(data.frame(x = norm_data), aes(sample = x)) +
  # stat_qq_band(distribution = "norm", alpha = 0.3) +
  stat_qq_line(distribution = "norm") +
  stat_qq_point(distribution = "norm") +
  geom_abline(intercept = mean(norm_data), slope = sd(t_data), color = "orange") +
  geom_abline(intercept = median(norm_data), slope = mad(t_data), color = "blue") +
  geom_abline(intercept = median(norm_data), slope = robustbase::Qn(t_data), color = "green") +
  geom_abline(intercept = median(norm_data), slope = robustbase::Sn(t_data), color = "red") +
  xlab("Theoretical quantiles") +
  xlab("Sample quantiles")

ggplot(data.frame(x = chisq1), aes(sample = x)) +
  # stat_qq_band(distribution = "norm", alpha = 0.3) +
  stat_qq_line(distribution = "norm") +
  stat_qq_point(distribution = "norm") +
  geom_abline(intercept = mean(chisq1), slope = sd(chisq1), color = "orange") +
  geom_abline(intercept = median(chisq1), slope = mad(chisq1), color = "blue") +
  geom_abline(intercept = median(chisq1), slope = robustbase::Qn(chisq1), color = "green") +
  geom_abline(intercept = median(chisq1), slope = robustbase::Sn(chisq1), color = "red") +
  xlab("Theoretical quantiles") +
  xlab("Sample quantiles")
```




## Discussion

This paper presents the \pkg{qqplotr} package, an extension of \pkg{ggplot2} that implements Q-Q plots in both the standard and de-trended orientations, reference lines for Q-Q plots, and confidence bands for Q-Q plots. The examples illustrate how to create Q-Q plots for non-standard distributions found outside of the \pkg{stats}, how to create de-trend Q-Q plots, and how to create Q-Q plots when data are grouped. Further, in the BRFSS example, we illustrated how jittering can be used in Q-Q plots to better compare discretized data to a continuous distribution.


Q-Q plots are members of the larger probability plotting family, and future versions of \pkg{qqplotr} will incorporate additional plots. Specifically, two types of probability plots will be included:

1. A standardarized P-P plot is constructed for location-scale distributions by plotting $F(z_{(i)})$ against $p_i$, where $z_{(i)}$ denotes the ordered and standardized observations and $p_i$ is denotes the plotting position \citep{Gan1991-yk}. As \citet{Gan1991-yk} point out, the probability plotting positions are equally spaced on the $x$-axis, whereas points on the $x$-axis of a Q-Q plot are more concentrated in the higher-density regions of the reference distribution. This leads to Q-Q plots being more sensitive to discrepencies in the tails of the distribution and standardized P-P plots being more sensitive to discrepencies in the "middle" of the proposed distribution. 
2. In engineering applications, probability plots refer to a linearized plot of $F(x_i)$ against $x_i$. The linearizing transformation of the hypothesized CDF is found by first finding a transformation that linearizes the quantile function---i.e., linearizing the plot of $F^{-1}(p)$ vs $x_p$---and then mapping $F^{-1}(p)$ back to the probability scale \citep[cf.][Chapter 6]{Meeker1998}.

It is important to note that all of the probability plotting methods discussed----Q-Q plots, standardized P-P plots, and probability plots---are invariant to linear transformations. Consequently, each type of plot will exhibit a linear relationship between the sample and hypothesized distribution for location-scale families, even if the location and/or scale parameters in the sample differ from the hypothesized distribution. This is not the case with unstandardized P-P plots \citep{Wilk1968-ii}, so they will not be included in \pkg{qqplotr}.


We have made design choices in \pkg{qqplotr} that we believe are in line with best practices for distributional assessment, but the implementation is flexible enough to allow for easy customization. For example, maximum likelihood estimation is used to estimate the parameters of the proposed model, but if outliers are present robust estimators may be desirable, such as if you are comparing the empirical distribution to the normal distribution. In this scenario, robust estimates of the location and scale can be obtained using the \pkg{robustbase} package \citep{robustbase}, and specified using the `dparams` argument directly. This is especially useful if you wish to use the parametric bootstrap to build confidence bands. Similarly, `stat_qq_line` only implements a reference line that passes through the first and third quartiles of the distributions. While this is the most commonly used reference line, alternative reference lines can be quickly implemented using `ggplot2::geom_abline` by specifying the slope and intercept.







<!-- For Q-Q plots, a reference line is often drawn through the points $(q(.25), q_y(.25))$ and $(q(.75), q_y(.75))$. -->

<!-- If the two distributions are identical, then this scatterplot will be linear with slope 1 and intercept 0. If the two distributions differ only by a linear transformation---such as $N(0, 1)$ and $N(10, 5)$---a Q-Q plot comparing the distributions will be linear, but with a different slope and intercept. -->

<!-- Probability plotting refers to a family of methods based on the cumulative distribution function (CDF), most notably quantile (Q-Q) plots and probability (P-P) plots \citep{Wilk1968-ii}. In this paper, we focus on comparing an empirical distribution to a theoretical distribution. Let $Y_1, \ldots, Y_n$ denote a random sample from an unknown population, and let $\widehat{F}_y(q)$ be the empirical cumulative distribution obtained from the sample. Further, let $F(q)$ denote the CDF of a proposed dsitribution for the sample. A Q-Q plot is constructed by plotting the quantiles of the empirical distribution, $q_y(p) = F_y^{-1}(p)$, against the corresponding quantiles of the theoretical distribution, $q(p) = F^{-1}(p)$. This constuction is illustrated in Figure \ref{fig:qq}. -->
<!-- A P-P plot is constructed by plotting $F(q)$ against $\widehat{F}_y(q)$ for various quantiles, $q$. This constuction is illustrated in Figure \ref{fig:pp}. Regardless of the plot constructed, if the two distributions are identical, then the scatterplots will be linear with slope 1 and intercept 0. Additionally, Q-Q plots are invariant to linear transformations, so if two random variables differ by a linear transformation a Q-Q plot showing draws from their distributions will still be linear, but with a different slope and intercept, as seen in Figure \ref{fig:qq}. P-P plots, in turn, are sensitive to linear transformations. -->


<!-- Innovations to Q-Q and P-P plots have also been proposed. \citet{Loy2016-fg} discuss the creation of detrended Q-Q plots, where the $y$-axis is changed to show the difference between $q_y$ and the reference line. Consequently, the line representing the agreement with the theoretical distribution is the $x$-axis. \citet{Loy2016-fg} find that detrended Q-Q plots are more powerful than other designs, so long at the $y$-axis limits are set so that the aspect ratio is kept the same as in the traditional Q-Q plot. In reliability and survival analysis, probability plots often refer to a hybrid probability plot, there the CDF of the proposed theoretical distribution is plotted against the empirical order statistics, and transformations are applied to each axis to linearize the CDF \citep[cf.][chapter 6]{Meeker1998}. This hybrid probability plot is invariant to linear transformations. -->


<!-- ```{r, echo=FALSE, out.width='.31\\linewidth', fig.width=3.5, fig.height=3.5, fig.show='hold', fig.align='center', fig.cap="\\label{fig:pp-designs}Illustrating different designs of probability plots."} -->
<!-- rand_sample$ppoints <- ppoints(rand_sample$rs) -->
<!-- ggplot(data = rand_sample, mapping = aes(x = sort(rs), y = ppoints)) + -->
<!--   geom_point() + -->
<!--   xlab("Sample quantiles") + -->
<!--   ylab("Cumulative probabilty") + -->
<!--   theme_minimal() + -->
<!--   ggtitle("Hybrid probability plot") -->
<!-- ``` -->

\bibliography{RJreferences}

## Acknowledgements

This work was partially funded by Google Summer of Code 2017.
